A Parallel General Purpose Multi-Objective Optimization Framework, with Application To Beam Dynamics

### 2018-07-11 paper/prab.tex
reformatting for prab

\begin{figure}%[h!]
%\begin{center}\normalsize
	\scalebox{0.8}{
	\pgfdeclarelayer{background}
	\pgfsetlayers{background,main}
	\begin{tikzpicture}[text=black]%[scale=0.05]%, transform shape]
	\input{figures/opt-framework_code.tex}
	\end{tikzpicture}
}
%\end{center}
%\end{center}
\caption{Multi-objective framework: the pilot (master) solves the
	optimization problem specified in the input file by coordinating optimizer
	algorithms and workers running forward solves.}
\end{figure}\label{fig:framenetwork}

@@

\begin{figure}
    \centering
    \begin{tikzpicture}[scale=0.8, transform shape, text=black]
      \input{figures/ga_explained_big.tex}
    \end{tikzpicture}
  \caption{Schematic view of interplay between selector and variator. The
  selector ranks all individuals in the population according to fitness and
  subsequently the variator uses the fittest individuals to produces new
  offspring. Finally, the new children are reintroduced in the population.}
  \label{fig:varsel}
\end{figure}

@@

The basic assumption in simulation-based optimization is that we need to
  call an expensive simulation software component present in the
  constraints or objectives.
We divide the framework in three exchangeable components, as shown in
  Figure~\ref{fig:opt-framework-layout}, to encapsulate the major behavioral
  patterns of the framework.
%
\begin{figure}
  \centering
  \begin{tikzpicture}[text=black]
    \input{figures/opt-framework-layout.tex}
  \end{tikzpicture}

@@

\noindent\begin{minipage}{\textwidth}
\begin{code}[Assembling the optimizer from components]
typedef OpalInputFileParser Input_t;
typedef PisaVariator        Opt_t;
typedef OpalSimulation      Sim_t;

typedef Pilot< Input_t, Opt_t, Sim_t > Pilot_t;
\end{code}
\end{minipage}

@@

\noindent\begin{minipage}{\textwidth}
\begin{code}[Optimizer API]
virtual void initialize() = 0;

// Poller hooks
virtual void setupPoll() = 0;
virtual void prePoll() = 0;
virtual void postPoll() = 0;
virtual void onStop() = 0;
virtual bool onMessage(MPI_Status status,
                       size_t length) = 0;
\end{code}
\end{minipage}

@@

\noindent\begin{minipage}{\textwidth}
\begin{code}[Creating function pointer for registering functor]
functionDictionary_t funcs;
client::function::type ff;
ff = average();
funcs.insert(std::pair<std::string, client::function::type>
        ("my_average_name", ff));
\end{code}
\end{minipage}

@@

If the objectives and constraints are simple arithmetical expressions, the
  \texttt{FunctionEvaluator} simulator can be used.
Using functors and the default expression primitives already powerful
  multi-objective optimization problems can be specified, i.e.\ the benchmark
  problem presented in \cite{hbwh:05}:
%
\begin{align} \label{eqn:bench}
  \text{ min} \quad & \quad \left[ 1 - \exp \left( -1 \left(
    \left(x_1 - \frac{1}{\sqrt{3}} \right)^2 +
    \left(x_2 - \frac{1}{\sqrt{3}} \right)^2 +
    \left(x_3 - \frac{1}{\sqrt{3}} \right)^2 \right)\right), \right. \\
                    & \quad \left. 1 - \exp \left( -1 \left(
    \left(x_1 + \frac{1}{\sqrt{3}} \right)^2 +
    \left(x_2 + \frac{1}{\sqrt{3}} \right)^2 +
    \left(x_3 + \frac{1}{\sqrt{3}} \right)^2 \right)\right) \right]^T \nonumber \\
    \text{s.t.} \quad & \quad -1 \le x_i \le 1, \quad i=1,2,3 \nonumber
  \text{.}
\end{align}
%
Using the default expression primitives this can be stated in an input file
  as:

>>>>

\begin{figure}%[h!]
%\begin{center}%\normalsize
	\scalebox{0.8}{
	\pgfdeclarelayer{background}
	\pgfsetlayers{background,main}
	\begin{tikzpicture}[text=black]%[scale=0.05]%, transform shape]
	\input{figures/opt-framework_code.tex}
	\end{tikzpicture}
}
%\end{center}
\caption{Multi-objective framework: the pilot (master) solves the
	optimization problem specified in the input file by coordinating optimizer
	algorithms and workers running forward solves.}
\end{figure}\label{fig:framenetwork}

@@

\begin{figure}
    \centering
    \scalebox{1}{
    \begin{tikzpicture}[scale=0.8, transform shape, text=black]
    \input{figures/ga_explained_big.tex}
    \end{tikzpicture}
	}
  \caption{Schematic view of interplay between selector and variator. The
  selector ranks all individuals in the population according to fitness and
  subsequently the variator uses the fittest individuals to produces new
  offspring. Finally, the new children are reintroduced in the population.}
  \label{fig:varsel}
\end{figure}

@@

The basic assumption in simulation-based optimization is that we need to
  call an expensive simulation software component present in the
  constraints or objectives.
We divide the framework in three exchangeable components, as shown in
  Figure~\ref{fig:opt-framework-layout}, to encapsulate the major behavioral
  patterns of the framework.
%
\begin{figure}
  \centering
  \scalebox{0.85}{
  \begin{tikzpicture}[text=black]
  \input{figures/opt-framework-layout.tex}
  \end{tikzpicture}
}

@@

%\noindent\begin{minipage}{0.48\textwidth}
%\noindent
\begin{code}[Assembling the optimizer \\ from components]
typedef OpalInputFileParser Input_t;
typedef PisaVariator        Opt_t;
typedef OpalSimulation      Sim_t;

typedef Pilot< Input_t, Opt_t, Sim_t > Pilot_t;
\end{code}
%\end{minipage}

@@

%\noindent\begin{minipage}{0.48\textwidth}
\begin{code}[Optimizer API]
virtual void initialize() = 0;

// Poller hooks
virtual void setupPoll() = 0;
virtual void prePoll() = 0;
virtual void postPoll() = 0;
virtual void onStop() = 0;
virtual bool onMessage(MPI_Status status,
                       size_t length) = 0;
\end{code}
%\end{minipage}

@@

%\noindent\begin{minipage}{\textwidth}
\begin{code}[Creating function pointer for registering functor]
functionDictionary_t funcs;
client::function::type ff;
ff = average();
funcs.insert(std::pair<std::string, 
		client::function::type> 
       		("my_average_name", ff));
\end{code}
%\end{minipage}

@@

If the objectives and constraints are simple arithmetical expressions, the
  \texttt{FunctionEvaluator} simulator can be used.
Using functors and the default expression primitives already powerful
  multi-objective optimization problems can be specified, i.e.\ the benchmark
  problem presented in \cite{hbwh:05}:
%
\small{
\begin{align} \label{eqn:bench}
  \text{ min} & \left[ 1 - \exp \left( -1 \left(
    \left(x_1 - \frac{1}{\sqrt{3}} \right)^2 +
    \left(x_2 - \frac{1}{\sqrt{3}} \right)^2 +
    \left(x_3 - \frac{1}{\sqrt{3}} \right)^2 \right)\right), \right. \\
                    & \quad \left. 1 - \exp \left( -1 \left(
    \left(x_1 + \frac{1}{\sqrt{3}} \right)^2 +
    \left(x_2 + \frac{1}{\sqrt{3}} \right)^2 +
    \left(x_3 + \frac{1}{\sqrt{3}} \right)^2 \right)\right) \right]^T \nonumber \\
    \text{s.t.} \quad & \quad -1 \le x_i \le 1, \quad i=1,2,3 \nonumber
  \text{.}
\end{align}
}
%
Using the default expression primitives this can be stated in an input file
  as:

### 2018-07-11 paper/theor-experiment.tex
reformatting for prab

\begin{figure}
	\begin{tikzpicture}[every node/.style={anchor=south west,inner sep=0pt},x=1mm, y=1mm,]   
	\node (fig2) at (0,0)
	{\includegraphics[width=\textwidth]{Report/tba_dogleg}};
	\node[fill=white, inner sep=2pt] (txt2) at (2,35) {$s_1$};
	\node[fill=white, inner sep=2pt] (txt2) at (14,35) {$s_2$};
	\node[fill=white, inner sep=2pt] (txt2) at (33,31) {$s_3$};
	\node[fill=white, inner sep=2pt] (txt2) at (24,50) {Septum};
	\draw [blue, ->, line width=1] (30,50) -- (30, 42);	
	\node[fill=white, inner sep=2pt] (txt2) at (3,50) {Kicker};
	\draw [blue, ->, line width=1] (9.5,50) -- (9.5, 43);
	\end{tikzpicture}
	\caption{Continuation of the high charge beam line layout at the AWA, top view. 
		This is the proposed two beam acceleration section.}
	\label{awa-tba}
\end{figure}
%
%\vspace{-1em}
In addition to the physics challenges, we chose this model to demonstrate the ability of the framework to tackle large problems.
Six design variables and objectives were used, along with three constraints.
The objectives include transverse and longitudinal beam sizes, 
transverse momentum, and longitudinal energy spread. 
The design variables include the two gun solenoids and 
the first four quadrupoles. The objectives are calculated at 
$s_3=19.4$~[m], at the entrance of the fifth quadrupole in Fig.~\ref{awa-tba}.  
This problem encompasses high dimensionality 
and nonlinear effects such as space charge. There is no existing information
that accurately predicts where the Pareto front will fall. This work is also meaningful in that
it will guide future operations at the AWA.
\nrnote{Moved final run set up and results to after time step scan and hyper parameters}

@@

\begin{table}%[h!]
	\begin{center}
		\caption{Checkmarks (\cmark) indicate desired physics is resolved. 
			An (\xmark) indicates the time step is too large.}
		\label{timestep}
		\begin{tabular*}{0.6\textwidth}{c c c c }
			\hline\noalign{\smallskip}
			Time Step (dT) & Linac & Drift & Quadrupoles \\
			\noalign{\smallskip}\hline\noalign{\smallskip}
			$5 \times10^{-10}$  & \xmark & \xmark & \xmark \\
			$1 \times10^{-10}$  & \xmark & \xmark & \xmark \\
			$5 \times10^{-11}$  & \xmark & \xmark & \xmark \\
			$1 \times10^{-11}$  & \cmark & \cmark & \xmark \\
			$5 \times10^{-12}$  & \cmark & \cmark & \xmark \\
			$1 \times10^{-12}$  & \cmark & \cmark & \cmark \\
			\noalign{\smallskip}\hline
		\end{tabular*}
	\end{center}
\end{table}

\iffalse
\begin{table}%[h!]
	\begin{center}
		\caption{Low and mid fidelity simulation parameters.}
		\label{fidelity}
		\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} C c D }
			\hline\noalign{\smallskip}
			& Number of particles & dT (seconds) & Time to simulation (minutes)\\
			\noalign{\smallskip}\hline\noalign{\smallskip}
			low fidelity  			&  10,000   & $5 \times10^{-11}$  &  2.6 \\
			adjusted low fidelity  	&  10,000   & $2 \times 10^{-12}$, $1 \times 10^{-11}$ & 1.6 \\
			mid fidelity 			&  100,000  & $1 \times 10^{-12}$ &  18\\
			\noalign{\smallskip}\hline
		\end{tabular*}
	\end{center}
\end{table}
\fi

\subsubsection{Hyper parameter Scan}
While the optimization problem and goals were well defined (Section \ref{awaproblem}), 
it was not clear what the best input parameters for the NSGA-II should be.
These parameters include gene mutation probability, mutation probability, 
recombination probability, number of individuals, 
and number of generations to complete. 
Given the beam line and design variables given in Section \ref{awaproblem},
\nrnote{Changed variables for final run, the variables in 5.3 no longer match these small runs. Need to rework this section} 
four optimization experiments were done with varying NSGA-II parameters. 
Similar to the time step scan, 
the goal of this exercise was to determine which set of optimization
parameters strongly influence the results, 
and whether there was a time to solution difference.
From here on, we will reference each experiment as ex-1, ex-2, ex-3, and ex-4
as shown in Table \ref{extable}. 
\begin{table}%[h!]
	\begin{center}
		\caption{Input Parameters for initial twenty four hour AWA optimization experiments. 
			The gene mutation probability was equal to the mutation probability (not shown) in all four experiments. 
			The max number of individuals per generation was~80.}
		\label{extable}
		\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} C C D }
			\hline\noalign{\smallskip}
			& Gene Mutation Probability & Recombination Probibility & Number of completed generations \\
			\noalign{\smallskip}\hline\noalign{\smallskip}
			ex-1 &  0.1  & 0.9  &  96 \\
			ex-2 &  0.3  & 0.7  &  81 \\
			ex-3 &  0.8  & 0.2  &  53 \\
			ex-4 &  0.01 & 0.09 &  95 \\
			\noalign{\smallskip}\hline
		\end{tabular*}
	\end{center}
\end{table}

@@

\begin{figure}
	\centering
	\begin{subfigure}{0.49\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Report/ex-pareto1}
		\caption{Pareto fronts at $s_1$.}
		\label{expareto1}
	\end{subfigure}
	\begin{subfigure}{0.49\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Report/ex-pareto2}
		\caption{Pareto fronts at $s_2$.}
		\label{expareto2}
	\end{subfigure}
	\caption{Comparison of Pareto fronts for initial optimization experiments.}
	\label{expareto}
\end{figure}

\subsubsection{AWA Optimization Problem}
\begin{align}
\text{min}  \quad & \text{rms}_{x}\left(s = s_3\right), \quad \text{rms}_{y}\left(s = s_3\right) \label{eq:awa:p1}\\
& \text{rms}_{px}\left(s = s_3\right), \quad \text{rms}_{py}\left(s = s_3\right), \label{eq:awa:p2}\\
& \text{rms}_{s}\left(s = s_3\right), \quad dE \label{eq:awa:p4} \\
\text{subject to} \quad & rms_x(s=s_1) < 0.1 \, [m]\label{eq:awa:c1}\\
\quad & rms_y(s=s_1) < 0.1\, [m]\label{eq:awa:c2}\\
\quad & |rms_y(s=s_1) - rms_x(s=s_1) | < 0.005 \, [m]\label{eq:awa:c3}\\
\quad & q = 40 \left[\text{nC}\right] \label{eq:awa:firstconstr}\\
\quad & \text{Volt}_{\text{Gun}} = 64\left[\text{MV/m}\right] \label{eq:awa:lastconstr}\\
\quad & \text{Volt}_{\text{Linac}} = 24-25\left[\text{MV/m}\right] \\
\quad & R_x = R_y = 9 \left[\text{mm}\right] \label{eq:awa:firstdvar}\\
\quad & \phi_{\text{gun}} =-20^\circ \label{eq:awa:gphidvar}\\
\quad & \phi_{\text{linac}} =-20^\circ \label{eq:awa:lastdvar}
\end{align}

>>>>

\begin{figure}
	\begin{tikzpicture}[every node/.style={anchor=south west,inner sep=0pt},x=1mm, y=1mm,]   
	\node (fig2) at (0,0)
	{\includegraphics[width=\textwidth]{Report/tba_dogleg}};
	\node[fill=white, inner sep=2pt] (txt2) at (2,35) {$s_1$};
	\node[fill=white, inner sep=2pt] (txt2) at (14,35) {$s_2$};
	\node[fill=white, inner sep=2pt] (txt2) at (33,31) {$s_3$};
	\node[fill=white, inner sep=2pt] (txt2) at (24,50) {Septum};
	\draw [blue, ->, line width=1] (30,50) -- (30, 42);	
	\node[fill=white, inner sep=2pt] (txt2) at (3,50) {Kicker};
	\draw [blue, ->, line width=1] (9.5,50) -- (9.5, 43);
	\end{tikzpicture}
	\caption{Continuation of the high charge beam line layout at the AWA, top view. 
		This is the proposed two beam acceleration section.}
	\label{awa-tba}
\end{figure}
%
%\vspace{-1em}

In addition to the physics challenges, we chose this model to demonstrate the ability of the framework to tackle large problems.
Six design variables and objectives were used, along with three constraints.
The objectives include transverse and longitudinal beam sizes, 
transverse momentum, and longitudinal energy spread. 
The design variables include the two gun solenoids and 
the first four quadrupoles. The objectives are calculated at 
$s_3=19.4$~[m], at the entrance of the fifth quadrupole in Fig.~\ref{awa-tba}.  
This problem encompasses high dimensionality 
and nonlinear effects such as space charge. There is no existing information
that accurately predicts where the Pareto front will fall. This work is also meaningful in that
it will guide future operations at the AWA.
\nrnote{Moved final run set up and results to after time step scan and hyper parameters}

@@

\begin{table}%[h!]
	\begin{center}
		\caption{Checkmarks (\cmark) indicate desired physics is resolved at that time step. 
			An (\xmark) indicates the time step is too large, and results are nonphysical.}
		\label{timestep}
		\begin{tabular*}{0.48\textwidth}{c @{\extracolsep{\fill}} C c D }
			\hline\noalign{\smallskip}
			Time Step (dT) & Linac & Drift & Quadrupoles \\
			\noalign{\smallskip}\hline\noalign{\smallskip}
			$5 \times10^{-10}$  & \xmark & \xmark & \xmark \\
			$1 \times10^{-10}$  & \xmark & \xmark & \xmark \\
			$5 \times10^{-11}$  & \xmark & \xmark & \xmark \\
			$1 \times10^{-11}$  & \cmark & \cmark & \xmark \\
			$5 \times10^{-12}$  & \cmark & \cmark & \xmark \\
			$1 \times10^{-12}$  & \cmark & \cmark & \cmark \\
			\noalign{\smallskip}\hline
		\end{tabular*}
	\end{center}
\end{table}

\iffalse
\begin{table}%[h!]
	\begin{center}
		\caption{Low and mid fidelity simulation parameters.}
		\label{fidelity}
		\begin{tabular*}{0.48\textwidth}{l @{\extracolsep{\fill}} C c D }
			\hline\noalign{\smallskip}
			& Number of particles & dT (seconds) & Time to simulation (minutes)\\
			\noalign{\smallskip}\hline\noalign{\smallskip}
			low fidelity  			&  10,000   & $5 \times10^{-11}$  &  2.6 \\
			adjusted low fidelity  	&  10,000   & $2 \times 10^{-12}$, $1 \times 10^{-11}$ & 1.6 \\
			mid fidelity 			&  100,000  & $1 \times 10^{-12}$ &  18\\
			\noalign{\smallskip}\hline
		\end{tabular*}
	\end{center}
\end{table}
\fi

\subsubsection{Hyper parameter Scan}
While the optimization problem and goals were well defined (Section \ref{awaproblem}), 
it was not clear what the best input parameters for the NSGA-II should be.
These parameters include gene mutation probability, mutation probability, 
recombination probability, number of individuals, 
and number of generations to complete. 
Given the beam line and design variables given in Section \ref{awaproblem},
\nrnote{Changed variables for final run, the variables in 5.3 no longer match these small runs. Need to rework this section} 
four optimization experiments were done with varying NSGA-II parameters. 
Similar to the time step scan, 
the goal of this exercise was to determine which set of optimization
parameters strongly influence the results, 
and whether there was a time to solution difference.
From here on, we will reference each experiment as ex-1, ex-2, ex-3, and ex-4
as shown in Table \ref{extable}. 
\begin{table}%[h!]
	\begin{center}
		\caption{Input Parameters for initial twenty four hour AWA optimization experiments. 
			The gene mutation probability was equal to the mutation probability (not shown) in all four experiments. 
			The max number of individuals per generation was~80.}
		\label{extable}
		\begin{tabular*}{0.48\textwidth}{l @{\extracolsep{\fill}} C C D }
			\hline\noalign{\smallskip}
			& Gene Mutation Probability & Recombination Probibility & Number of completed generations \\
			\noalign{\smallskip}\hline\noalign{\smallskip}
			ex-1 &  0.1  & 0.9  &  96 \\
			ex-2 &  0.3  & 0.7  &  81 \\
			ex-3 &  0.8  & 0.2  &  53 \\
			ex-4 &  0.01 & 0.09 &  95 \\
			\noalign{\smallskip}\hline
		\end{tabular*}
	\end{center}
\end{table}

@@

\begin{figure}
	\centering
	\begin{minipage}{0.49\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Report/ex-pareto1}
		(a) Pareto fronts for ex-1 through ex-4 at $s_1$.
		%\caption{(a) Pareto fronts at $s_1$.}
		%\label{expareto1}
	\end{minipage}
	\begin{minipage}{0.49\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Report/ex-pareto2}
		(b) Pareto fronts for ex-1 through ex-4 at $s_2$.
		%\caption{(b) Pareto fronts at $s_2$.}
		%\label{expareto2}
	\end{minipage}
	\caption{Comparison of Pareto fronts for initial optimization experiments, ex-1 through ex-4.}
	\label{expareto}
\end{figure}

\subsubsection{TBA Optimization Problem}
\begin{align}
\text{min}  \quad & \text{rms}_{x}\left(s = s_3\right), \quad \text{rms}_{y}\left(s = s_3\right) \label{eq:awa:p1}\\
& \text{rms}_{px}\left(s = s_3\right), \quad \text{rms}_{py}\left(s = s_3\right), \label{eq:awa:p2}\\
& \text{rms}_{s}\left(s = s_3\right), \quad dE \label{eq:awa:p4} \\
\text{subject to} \quad & rms_x(s=s_1) < 0.1 \, [m]\label{eq:awa:c1}\\
\quad & rms_y(s=s_1) < 0.1\, [m]\label{eq:awa:c2}\\
\quad & |rms_y(s=s_1) - rms_x(s=s_1) | < 0.005 \, [m]\label{eq:awa:c3}\\
\quad & q = 40 \left[\text{nC}\right] \label{eq:awa:firstconstr}\\
\quad & \text{Volt}_{\text{Gun}} = 64\left[\text{MV/m}\right] \label{eq:awa:lastconstr}\\
\quad & \text{Volt}_{\text{Linac}} = 24-25\left[\text{MV/m}\right] \\
\quad & R_x = R_y = 9 \left[\text{mm}\right] \label{eq:awa:firstdvar}\\
\quad & \phi_{\text{gun}} =-20^\circ \label{eq:awa:gphidvar}\\
\quad & \phi_{\text{linac}} =-20^\circ \label{eq:awa:lastdvar}
\end{align}
